From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/applications.tex
Progress

\[
  \begin{tikzpicture}[->,>=stealth',auto,thick,node distance=1em]
    % nodes
    \node (A) at (0, 2) {\Afun{reversibleOr1}};
    \node (B) at (0, 0) {\Afun{reversibleOr2}};
    \node (C) at (4, 1) {\Afun{reversibleOrNorm}};
    \node (D) at (9, 1) {\AgdaSymbol{(}\AgdaNumber{0 5 6 7 4 1 2 3}\AgdaSymbol{)}};
    % arrows
    \path
    (A) edge[bend left] node [above left] { $\strut\normt_{1}$ } (C)
    (B) edge[bend right] node [below left] { $\strut\normt_{1}$ } (C)
    (D) edge node[above] { $\strut\term{quote}$ } (C);
  \end{tikzpicture}
\]

@@

As an example, consider the following problem. Toffoli gate on 3 bits can be extended to 4 bits, inserting a wire in
four possible positions. The function of such gate is then identity on the newly inserted bit, and controlled-controlled
not on the three remaining ones.

\medskip
\extendedToffoli3{}

On the other hand, it is possible to construct a proper 4-bit toffoli gate:

\medskip
\toffoli4{}

Now, the question we want to asnwer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
behaviour of a 4-bit toffoli gate?

@@

In our case, we define the $\term{parity}$ function:
parity : {n m : ℕ} → (c : S n Pi^.⟷₁^ S m) → ⊤ ⊔ ⊤
parity {X} {Y} c = list-len-parity (pi^2list c)

Thus, we formulate the theorem using the representation on lists of transpositions, and transfer it to circuits using
the equivalences proven in the previous sections.

>>>>

% \[
%   \begin{tikzpicture}[->,>=stealth',auto,thick,node distance=1em]
%     % nodes
%     \node (A) at (0, 2) {\Afun{reversibleOr1}};
%     \node (B) at (0, 0) {\Afun{reversibleOr2}};
%     \node (C) at (4, 1) {\Afun{reversibleOrNorm}};
%     \node (D) at (9, 1) {\AgdaSymbol{(}\AgdaNumber{0 5 6 7 4 1 2 3}\AgdaSymbol{)}};
%     % arrows
%     \path
%     (A) edge[bend left] node [above left] { $\strut\normt_{1}$ } (C)
%     (B) edge[bend right] node [below left] { $\strut\normt_{1}$ } (C)
%     (D) edge node[above] { $\strut\term{quote}$ } (C);
%   \end{tikzpicture}
% \]

@@

As an example, consider the following problem. Toffoli gate on 3 bits can be extended to 4 bits, inserting a wire in
four possible positions. The function of such gate is then identity on the newly inserted bit, and controlled-controlled
not on the three remaining ones.
\medskip
\extendedToffoli{}
\baselineskip
On the other hand, it is possible to construct a proper 4-bit toffoli gate:
\medskip
\toffoli{}
\baselineskip
Now, the question we want to answer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
behaviour of a 4-bit toffoli gate?

@@

We define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
$c$. It is implemented as the transformation from a $\PiLang$ program into a list of transposition, using $\evalh_{1}$
defined in~\cref{sec:equivalence}, and computing the parity on this representation, which means simply checking the
parity of the length of the list.

\begin{propositionrep}
  \label{prop:parity-preserved}
  For two $\PiLang$ circuits $c_1, c_2$, if there exists a 2-combinator between them, then $\parity c_1 = \term{parity} c_2$. 
\end{propositionrep}
\begin{proof}
  First, it is easy to check that Coxeter relations preserve parity of the list. Second, if $c_1, c_2$ are related by a
  2-combinator, then their list-of-transpositions representations are related in the Coxeter relation. Composing these
  two facts completes the proof.
\end{proof}

Using ~\cref{prop:parity-preserved} and showing that any composition of even cicuits yields an even ciruit, we can
answer the question by computing the parity of the \Afun{toffoli₄} and all of the \Afun{toffoli₃} gates.

Thus, we formulated the theorem using the representation on lists of transpositions, and transferred it to circuits using
the equivalences proven in the previous sections.

### 2021-10-30 Pi/paper/discussion.tex
Progress

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work. In the
remainder of this section, we discuss some broader related work.

% \begin{center}
%     \begin{tabular}{c c c}
%         \hline
%         $\PiLang$     & $\Sn$               & $\UFin$     \\
%         \hline
%         types         & Natural numbers     & Finite sets \\
%         \hline
%         1-combinators & Generators of $\Sn$ & 1-paths     \\
%         \hline
%         2-combinators & Relations of $\Sn$  & 2-paths     \\
%         \hline
%     \end{tabular}
% \end{center}

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work, as in~\cref{fig:table-summary}. In the
remainder of this section, we discuss some broader related work.

\begin{figure}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}
            \hline
            \rowcolor{lightgray} $\PiLang$     & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
            \hline
            types         & Natural numbers                & Finite sets & 0-cells \tabularnewline
            \hline
            1-combinators & Generators of $\Sn$            & Bijections  & 1-paths \tabularnewline
            \hline
            2-combinators & Relations of $\Sn$             & Homotopies  & 2-paths \tabularnewline
            \hline
        \end{tabular}
    \end{center}
    \caption{Summary of the established correspondences}
    \label{fig:table-summary}        
\end{figure}
